---
title: "Computational Statistics & Probability"
author: "Lab 2 - Linear Models"
date: "Fall 2025"
output: pdf_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
answer_key <- FALSE
library(rethinking)
```


# Learning Objectives

By the end of this lab, you will be able to:

- Specify and justify priors for linear regression parameters
- Use prior predictive simulation to check if priors are reasonable
- Fit linear models using `quap()` 
- Interpret regression coefficients in terms of associations
- Make predictions with uncertainty for new observations
- Use posterior predictive checks to assess model fit

# Introduction: Predicting Height from Weight

In this lab, we'll build a linear model to predict adult height from weight using the !Kung San census data.

**The Model:**
$$
\begin{aligned}
\text{height}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta \cdot (\text{weight}_i - \bar{\text{weight}}) \\
\alpha &\sim \text{Normal}(178, 20) \\
\beta &\sim \text{Log-Normal}(0, 1) \\
\sigma &\sim \text{Uniform}(0, 50)
\end{aligned}
$$

**Key concepts:**

  - $\alpha$ (alpha): Average height when weight = mean weight
  - $\beta$ (beta): Change in height per 1 kg increase in weight
  - $\sigma$ (sigma): Standard deviation around the line
  - Centering weight makes $\alpha$ interpretable and improves computation

```{r load-data}
# be sure `rethinking` is loaded
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ] # adults only  # Adults only

# Summary
cat("Number of adults:", nrow(d2), "\n")
cat("Weight range:", min(d2$weight), "to", max(d2$weight), "kg\n")
cat("Height range:", min(d2$height), "to", max(d2$height), "cm\n")
```


# 1. Prior Predictive Simulation

Before fitting, check if priors make sense.



**a)** Understanding the Priors


We choose:
- $\alpha \sim$ Normal(178, 20): Average height $\approx 178$ cm, $\pm 20$ cm uncertainty
- $\beta \sim$  Log-Normal(0, 1): Positive relationship (can't shrink by gaining weight!)
- $\sigma \sim$  Uniform(0, 50): Residual variation up to 50 cm

Why use Log-Normal for $\beta$ instead of Normal? (1-2 sentences)
```{r prior-explain, include=!answer_key}
# YOUR ANSWER:
```

```{r prior-explain-ans, include=answer_key}
# ANSWER: 
# Log-Normal is strictly positive, which makes biological sense:
# gaining weight shouldn't make you shorter. Normal would allow
# negative beta values (people getting shorter as they get heavier).
```


**b)** Simulate from Priors

Simulate N = 100 prior regression lines:
```{r prior-sim, include=!answer_key}
set.seed(212)
N <- 100

# Sample from priors
# alpha <- # TODO #YOUR CODE HERE
# beta <- # TODO #YOUR CODE HERE

# Plot
# plot(NULL, 
#      xlim = c(30, 70), ylim = c(50, 250),
#      xlab = "Weight (kg)", ylab = "Height (cm)",
#      main = "Prior Predictive Simulation")
# 
# abline(h = 0, lty = 2, col = "gray")
# abline(h = 272, lty = 2, col = "gray")
# 
# xbar <- mean(d2$weight)
# for (i in 1:N) {
#   curve(alpha[i] + beta[i] * (x - xbar),
#         from = 30, to = 70, add = TRUE,
#         col = col.alpha("black", 0.2))
# }
```
```{r prior-sim-ans, include=answer_key}
set.seed(212)
N <- 100

alpha <- rnorm(N, 178, 20)
beta <- rlnorm(N, 0, 1)

plot(NULL, 
     xlim = c(30, 70), ylim = c(50, 250),
     xlab = "Weight (kg)", ylab = "Height (cm)",
     main = "Prior Predictive Simulation")

abline(h = 0, lty = 2, col = "gray")
abline(h = 272, lty = 2, col = "gray")
text(35, 272, "Tallest human", pos = 3, cex = 0.8)

xbar <- mean(d2$weight)
for (i in 1:N) {
  curve(alpha[i] + beta[i] * (x - xbar),
        from = 30, to = 70, add = TRUE,
        col = col.alpha("black", 0.2))
}

# INTERPRETATION:
# Most lines positive (increasing height with weight)
# Most fall in plausible ranges (100-200 cm)
# Some extremes, but that's OK with vague priors
# No negative slopes (thanks to Log-Normal)
```

**c)** Check Prior Predictions
For 50 kg adult, what heights does prior predict?
```{r prior-check, include=!answer_key}
# predicted_heights <- # TODO (use alpha, beta from part b)
# mean(predicted_heights)
# PI(predicted_heights, prob = 0.89)
```

```{r prior-check-ans, include=answer_key}
weight_test <- 50
predicted_heights <- alpha + beta * (weight_test - xbar)

cat("For 50 kg adult, prior predicts:\n")
cat("  Mean:", round(mean(predicted_heights), 1), "cm\n")
cat("  89% PI:", round(PI(predicted_heights, 0.89), 1), "cm\n")

```


# 2. Fit the Model
```{r fit-model}
m4.3 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b * (weight - xbar),
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ), 
  data = d2
)

precis(m4.3)
```

*a)* Interpret Coefficients
```{r interpret, include=!answer_key}
# 1. What is posterior mean for intercept (a)? What does it mean?
# YOUR ANSWER:

# 2. What is posterior mean for slope (b)? What does it mean?
# YOUR ANSWER:

# 3. Has data updated beliefs about beta?
#    Prior Log-Normal(0,1) has mean approx. 1.65
# YOUR ANSWER:
```

```{r interpret-ans, include=answer_key}
# 1. Intercept (a): approx. 154.6 cm
#    Expected height at average weight (44.99 kg)
#    Centering makes this directly interpretable

# 2. Slope (b): approx. 0.90 cm/kg
#    Each 1 kg increase → ~0.90 cm taller on average
#    This is ASSOCIATION, not causation

# 3. Prior vs. Posterior:
#    Prior: mean = approx. 1.65, very wide
#    Posterior: mean = approx. 0.90, narrow (0.84-0.97)
#    Data substantially updated beliefs!

# 4. Sigma: approx. 5.1 cm
#    Even accounting for weight, heights vary approx. 5 cm
#    Other factors: genetics, nutrition, etc.
```

**b)** Prior vs. Posterior Comparison
```{r prior-post, include=answer_key}
post <- extract.samples(m4.3)

par(mfrow = c(1, 2))

# Prior for beta
prior_beta <- rlnorm(10000, 0, 1)
dens(prior_beta, xlim = c(0, 4),
     main = "Prior for Beta", xlab = "beta (cm/kg)")

# Posterior for beta
dens(post$b, xlim = c(0, 4),
     main = "Posterior for Beta", xlab = "beta (cm/kg)",
     col = "steelblue", lwd = 2)

par(mfrow = c(1, 1))

# Prior was vague (0 to 4+)
# Posterior concentrated around 0.9
# Bayesian learning!
```


# 3. Make Predictions

Five new adults - predict their heights:

| Individual | Weight (kg) | Predicted Height | 89% Interval |
|:----------:|:-----------:|:----------------:|:------------:|
| 1          | 47          |                  |              |
| 2          | 60          |                  |              |
| 3          | 37          |                  |              |
| 4          | 51          |                  |              |
| 5          | 43          |                  |              |

```{r predictions, include=!answer_key}
# new_data <- data.frame(
#   individual = 1:5,
#   weight = c(47, 60, 37, 51, 43)
# )

# height_sim <- # TODO: use sim()
# Exp_height <- # TODO: mean for each individual
# height_CI <- # TODO: 89% CI for each
```

```{r predictions-ans, include=answer_key}
new_data <- data.frame(
  individual = 1:5,
  weight = c(47, 60, 37, 51, 43)
)

height_sim <- sim(m4.3, data = new_data)
Exp_height <- apply(height_sim, 2, mean)
height_CI <- apply(height_sim, 2, PI, prob = 0.89)

new_data$predicted <- round(Exp_height, 1)
new_data$CI_lower <- round(height_CI[1, ], 1)
new_data$CI_upper <- round(height_CI[2, ], 1)

library(knitr)
kable(new_data, col.names = c("Individual", "Weight (kg)", 
      "Predicted (cm)", "89% Lower", "89% Upper"))
```

## Interpret Predictions
```{r interp-pred, include=!answer_key}
# For Individual 2 (60 kg):
# 1. Predicted height?
# 2. What does 89% interval tell you?
# 3. Why uncertainty even though we know weight?
```

```{r interp-pred-ans, include=answer_key}
# 1. approx. 168 cm
# 2. 89% confident true height in interval (approx. 158-178 cm)
#    Accounts for parameter uncertainty + individual variation
# 3. Two sources of uncertainty:
#    - Parameter: don't know alpha,beta,sigma exactly (epistemic)
#    - Individual: people vary at same weight (aleatoric)
#    This is why sim() not just link()
```

# 4. Visualize Regression Line
```{r reg-plot, include=!answer_key}
# weight_seq <- # TODO
# mu <- # TODO: use link()
# mu_mean <- # TODO
# mu_PI <- # TODO

# plot(height ~ weight, data = d2, col = col.alpha("black", 0.5))
# lines(weight_seq, mu_mean, col = "steelblue", lwd = 3)
# shade(mu_PI, weight_seq, col = col.alpha("steelblue", 0.3))
```

```{r reg-plot-ans, include=answer_key}
weight_seq <- seq(25, 70, by = 1)
mu <- link(m4.3, data = data.frame(weight = weight_seq))
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI, prob = 0.89)

plot(height ~ weight, data = d2, 
     col = col.alpha("black", 0.5),
     xlab = "Weight (kg)", ylab = "Height (cm)",
     main = "Height vs. Weight with 89% Uncertainty")

lines(weight_seq, mu_mean, col = "steelblue", lwd = 3)
shade(mu_PI, weight_seq, col = col.alpha("steelblue", 0.3))

legend("bottomright",
       legend = c("Data", "Mean", "89% interval"),
       pch = c(1, NA, 15),
       lty = c(NA, 1, NA),
       col = c("black", "steelblue", col.alpha("steelblue", 0.3)),
       lwd = c(NA, 3, NA))
```

# 5. Posterior Predictive Checks

**a)** Simulate for Existing Data
```{r ppc, include=!answer_key}
# height_post_pred <- # TODO: sim() for all d2
# pred_mean <- # TODO
# pred_PI <- # TODO
```

```{r ppc-ans, include=answer_key}
height_post_pred <- sim(m4.3, data = d2)

cat("Simulated dimensions:", dim(height_post_pred), "\n")

pred_mean <- apply(height_post_pred, 2, mean)
pred_PI <- apply(height_post_pred, 2, PI, prob = 0.89)

residuals <- d2$height - pred_mean
cat("Residual mean:", round(mean(residuals), 3), "\n")
cat("Residual SD:", round(sd(residuals), 2), "\n")
```

**b)** Visual Checks
```{r ppc-viz, eval=answer_key}
par(mfrow = c(1, 2))

# Observed vs. Predicted
plot(d2$height, pred_mean,
     xlab = "Observed (cm)", ylab = "Predicted (cm)",
     main = "Observed vs. Predicted",
     col = col.alpha("black", 0.5), pch = 16)
abline(0, 1, col = "red", lwd = 2, lty = 2)

# Residuals
plot(d2$weight, residuals,
     xlab = "Weight (kg)", ylab = "Residual (cm)",
     main = "Residuals vs. Weight",
     col = col.alpha("black", 0.5), pch = 16)
abline(h = 0, col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```


**c)** Check Coverage
```{r coverage, eval=answer_key}
in_interval <- (d2$height >= pred_PI[1, ]) & 
                (d2$height <= pred_PI[2, ])
coverage <- mean(in_interval)
cat("Coverage:", round(coverage, 3), "\n")
cat("Expected: 0.89\n")
if (abs(coverage - 0.89) < 0.05) {
  cat("Well-calibrated!\n")
} else {
  cat(" Coverage off\n")
}
```

# 6. Reflection Questions

**a)** Prior vs. Posterior Predictive

What's the difference? Why do both?
```{r ref-a, include=!answer_key}
# YOUR ANSWER:
```
```{r ref-a-ans, eval=answer_key}
# Prior predictive: Check priors BEFORE data
# Shows what we'd expect if priors were true
# Posterior predictive: Validate model AFTER fitting
# Shows if model reproduces observed patterns
# Both needed: assumptions reasonable (prior) + model adequate (posterior)
```

**b)** Uncertainty in Predictions

Why are individual predictions (Q3) wider than regression line (Q4)?
```{r ref-b, include=!answer_key}
# YOUR ANSWER:
```
```{r ref-b-ans, eval=answer_key}
# Regression line (link): Only parameter uncertainty
# Individual predictions (sim): Parameter uncertainty + individual variation
# Even if we knew true line, people vary around it (genetics, etc.)
```

**c)** Association vs. Causation

Does gaining weight CAUSE increased height?
```{r ref-c, include=!answer_key}
# YOUR ANSWER:
```
```{r ref-c-ans, include=answer_key}
# NO - this shows ASSOCIATION not causation
# Both influenced by common causes (genetics, age, nutrition)
# Could be reverse (tall -> heavy) or confounding
# Need experiments or causal models for causation!
```



# Summary

Complete Bayesian workflow for linear regression:

1.  Specified priors
2.  Prior predictive simulation
3.  Fit model with quap()
4.  Interpreted coefficients
5.  Made predictions with uncertainty
6.  Posterior predictive checks

**Key:** Always check before (prior) and after (posterior)!